Unshuffling Permutations

### 2015-09-22 submission.tex
Deplacement des demos dans submission.tex
abstract

\begin{abstract}
    A permutation is said to be a square if it can be obtained by
    shuffling two order-isomorphic patterns. The definition is intended
    to be the natural counterpart to the ordinary shuffle of words and
    languages. In this paper, we tackle the problem of recognizing square
    permutations from both the point of view of algebra and algorithmic.
    On the one hand, we present some algebraic and combinatorial
    properties of the shuffle product of permutations. We follow an
    unusual line consisting in defining the shuffle of permutations by
    means of an unshuffling operator, known as a coproduct. This
    strategy allows to obtain easy proofs for algebraic and combinatorial
    properties of our shuffle product. We besides exhibit a bijection
    between square $(213,231)$-avoiding permutations and square binary
    words. On the other hand, we prove that recognizing square
    permutations is \NPC. This property is proved by establishing a
    reduction from the pattern involvement problem and by using a
    containment-free perfect matching criterion to decide if a
    permutation is a square.
\end{abstract}

@@

The purpose of this section is to define a shuffle product $\SHUFFLE$
on permutations. To define this shuffle product, we shall first define
a coproduct denoted by $\Delta$, enabling to unshuffle permutations.
By duality, $\Delta$ implies the definition of $\SHUFFLE$. The reason
why we need to pass by the definition of $\Delta$ to define $\SHUFFLE$
is justified by the fact that a lot of properties of $\SHUFFLE$ depend
of properties of $\Delta$ and that this strategy allows to write concise
and clear proofs of them. We invite the reader unfamiliar with the
concepts of coproduct and duality to consult~\cite{Joni:Rota:1979}
or~\cite{Grinberg:Reiner:2014}.
\medskip

Let use denote by $\QQ[S]$ the linear span of all permutations. We
define a linear coproduct $\Delta$ on $\QQ[S]$ in the following way. For
any permutation $\pi$, we set
\begin{equation} \label{equ:unshuffling_coproduct}
    \Delta(\pi) =
    \sum_{P_1 \sqcup P_2 = [|\pi|]}
    \STD\left(\pi_{|P_1}\right) \otimes \STD\left(\pi_{|P_2}\right).
\end{equation}
We call $\Delta$ the {\em unshuffling coproduct of permutations}. For
instance,
\begin{equation}
    \Delta(213) =
    \epsilon \otimes 213 + 2. 1 \otimes 12 +
    1 \otimes 21 + 2. 12 \otimes 1 + 21 \otimes 1 + 213 \otimes \epsilon,
\end{equation}
\begin{equation}
    \Delta(1234) =
    \epsilon \otimes 1234 + 4. 1 \otimes 123 + 6. 12 \otimes 12 +
    4. 123 \otimes 1 + 1234 \otimes \epsilon,
\end{equation}
\begin{equation}\begin{split} \label{equ:example_unshuffling_coproduct}
    \Delta(1432) & =
    \epsilon \otimes 1432 + 3. {\bf 1 \otimes 132} + 1 \otimes 321 +
    3. 12 \otimes 21 \\ & + 3. 21 \otimes 12 + 3. 132 \otimes 1 +
    321 \otimes 1 + 1432 \otimes \epsilon.
\end{split}\end{equation}
Observe that the coefficient of the tensor $1 \otimes 132$ is $3$
in~\eqref{equ:example_unshuffling_coproduct} because there are exactly
three ways to extract from the permutation $1432$ two disjoint subwords
respectively order-isomorphic to the permutations $1$ and $132$.
\medskip

As announced, let us now use $\Delta$ to define a shuffle product on
permutations. As any coproduct, $\Delta$ leads to the definition of a
product obtained by duality in the following way.
From~\eqref{equ:unshuffling_coproduct}, for any permutation $\pi$, we
have
\begin{equation} \label{equ:schematic_coproduct}
    \Delta(\pi) =
    \sum_{\sigma, \nu \in S} \lambda_{\sigma, \nu}^\pi \;
    \sigma \otimes \nu,
\end{equation}
where the $\lambda_{\sigma, \nu}^\pi$ are nonnegative integers. Now,
by definition of duality, the dual product of $\Delta$, denoted by
$\SHUFFLE$, is a linear binary product on $\QQ[S]$. It satisfies, for
any permutations $\sigma$ and $\nu$,
\begin{equation}
    \sigma \SHUFFLE \nu =
    \sum_{\pi \in S}
    \lambda_{\sigma, \nu}^\pi \; \pi,
\end{equation}
where the coefficients $\lambda_{\sigma, \nu}^\pi$ are the ones
of~\eqref{equ:schematic_coproduct}. We call $\SHUFFLE$ the
{\em shuffle product of permutations}. For instance,
\begin{equation}\begin{split} \label{equ:example_shuffle_product}
    12 \SHUFFLE 21 & =
    1243 + 1324 + 2.1342 + 2.1423 + 3. {\bf 1432} + 2134 + 2.2314 \\
    & + 3.2341 + 2413 + 2.2431 + 2.3124 + 3142 + 3.3214 + 2.3241 \\
    & + 3421 + 3.4123 + 2.4132 + 2.4213 + 4231 + 4312.
\end{split}\end{equation}
Observe that the coefficient $3$ of the permutation $1432$
in~\eqref{equ:example_shuffle_product} comes from the fact that the
coefficient of the tensor $12 \otimes 21$ is $3$
in~\eqref{equ:example_unshuffling_coproduct}.
\medskip

Intuitively, this product shuffles the values and the positions of the
letters of the permutations. As first and easy properties of $\SHUFFLE$,
one can observe that the empty permutation $\epsilon$ is a unit for
$\SHUFFLE$ and that this product is graded by the sizes of the
permutations ({\em i.e.}, the product of a permutation of size $n$ with
a permutation of size $m$ produces a sum of permutations of size $n + m$).
Notice that there are other shuffle products on
permutations~\cite{DHT:IJAC:2002}: the {\em shifted shuffle} and the
{\em convolution product}. Our shuffle product is different from these
last two.
\medskip

We say that a permutation $\pi$ {\em appears} in the shuffle
$\sigma \SHUFFLE \nu$ of two permutations $\sigma$ and $\nu$ if the
coefficient $\lambda_{\sigma, \nu}^\pi$ defined above is different from
zero. In a more combinatorial way, this is equivalent to say that there
are two sets $P_1$ and $P_2$ of disjoints indexes of letters of $\pi$
satisfying $P_1 \sqcup P_2 = [|\pi|]$ such that the subword $\pi_{|P_1}$
is order-isomorphic to $\sigma$ and the subword $\pi_{|P_2}$ is
order-isomorphic to $\nu$.
\medskip

@@

\begin{tabular}{c|c|c}
    Size $0$ \, & Size $2$ & Size $4$ \\ \hline
    \multirow{2}{*}{\, $\epsilon$} &
    \multirow{2}{*}{\, $12$, $21$ \,} &
    $1234$, $1243$, $1423$, $1324$, $1342$, $4132$, $3124$, $3142$,
    $3412$, $4312$, \\
    & & $2134$, $2143$, $2413$, $4213$, $2314$, $2431$, $4231$, $3241$,
    $3421$, $4321$
\end{tabular}
\medskip

@@

%%
%% ---- Binary square words and permutations ----
%%
\section{Binary square words and permutations}
\label{section:Binary square words and permutations}
In this section, we shall show that the square binary words are in
one-to-one correspondence with square permutations avoiding some
patterns. This property establishes a link between the shuffle of binary
words and our shuffle of permutations and allows to obtain a new
description of square binary words.
\medskip

Let $u$ be a binary word of length $n$ with $k$ occurrences of $0$.
We denote by $\BINTOPERM$ the map sending any such word $u$ to the
permutation obtained by replacing from left to right each occurrence of
$0$ in $u$ by $1$, $2$, \dots, $k$, and from right to left each
occurrence of $1$ in $u$ by $k + 1$, $k + 2$, \dots, $n$. For instance,
\begin{equation}
    \BINTOPERM({\bf 1}00{\bf 1}0{\bf 1}{\bf 1}0{\bf 1}000) =
    {\bf C} 12 {\bf B} 3 {\bf A} {\bf 9} 4 {\bf 8} 567.
\end{equation}
Observe that for any nonempty permutation $\pi$ in the image of
$\BINTOPERM$, there is exactly one binary word $u$ such that
$\BINTOPERM(u0) = \BINTOPERM(u1) = \pi$. In support of this observation,
when $\pi$ has an even size, we denote by $\PERMTOBIN(\pi)$ the word $ua$
such that $|ua|_0$ and $|ua|_1$ are both even, where $a \in \{0, 1\}$.
\medskip

\begin{proposition} \label{prop:bijection_binary_to_permutations_squares}
    For any $n \geq 0$, the map $\BINTOPERM$ restricted to the set of
    square binary words of length $2n$ is a bijection between this last
    set and the set of square permutations of size $2n$ avoiding the
    patterns $213$ and $231$.
\end{proposition}
\begin{proof}
    [of Proposition~\ref{prop:bijection_binary_to_permutations_squares}]
    The statement of the proposition is a consequence of the
    following claims implying that $\PERMTOBIN$ is the inverse
    map of $\BINTOPERM$ over the set of square binary words.
    \begin{claim} \label{claim:binary_to_permutation_avoiding}
        The image of $\BINTOPERM$ is the set of all permutations
        avoiding $213$ and $231$.
    \end{claim}
    \begin{proof}[of Claim~\ref{claim:binary_to_permutation_avoiding}]
        Let us first show that the image of $\BINTOPERM$ contains only
        permutations avoiding $213$ and $231$. Let $u$ be a binary word,
        $\pi = \BINTOPERM(u)$, and $P_0$ (resp. $P_1$) be the set of the
        positions of the occurrences of $0$ (resp. $1$) in $u$. By
        definition of $\BINTOPERM$, from left to right, the subword
        $v = \pi_{|P_0}$ is increasing and the subword $w = \pi_{|P_1}$
        is decreasing, and all letters of $w$ are greater than those
        of~$v$. Now, assume that $\pi$ admits an occurrence of $213$.
        Then, since $v$ is increasing and $w$ is decreasing, there is an
        occurrence of $3$ (resp. $13$, $23$) in $v$ and a relative
        occurrence of $21$ (resp. $2$, $1$). All these three cases
        contradict the fact that all letters of $w$ are greater than
        those of $v$. A similar argument shows that $\pi$ avoids~$231$
        as well.
        \smallskip

        Finally, observe that any permutation $\pi$ avoiding $213$ and
        $231$ necessarily starts by the smallest possible letter or the
        greatest possible letter. This property is then true for the
        suffix of $\pi$ obtained by deleting its first letter,
        and so on for all of its suffixes. Thus, by
        replacing each letter $a$ of $\pi$ by $0$ (resp. $1$) if $a$ has the
        role of a smallest (resp. greatest) letter, one obtains a binary
        word $u$ such that $\BINTOPERM(u) = \pi$. Hence, all permutations
        avoiding $213$ and $231$ are in the image of $\BINTOPERM$.
        \qed
    \end{proof}
    \begin{claim} \label{claim:square_binary_to_square_permutation}
        If $u$ is a square binary word, $\BINTOPERM(u)$ is a square
        permutation.
    \end{claim}
    \begin{proof}[of Claim~\ref{claim:square_binary_to_square_permutation}]
        Since $u$ is a square binary word, there is a binary word $v$
        such that $u \in v \shuffle v$. Then, there are two disjoint
        sets $P$ and $Q$ of positions of letters of $u$ such that
        $u_{|P} = v = u_{|Q}$. Now, by definition of $\BINTOPERM$, the
        words $\BINTOPERM(u)_{|P}$ and $\BINTOPERM(u)_{|Q}$ have the
        same standarized $\sigma$. Hence, and by definition of
        the shuffle product of permutations, $\BINTOPERM(u)$ appears in
        $\sigma \SHUFFLE \sigma$, showing that $\BINTOPERM(u)$ is a
        square permutation.
        \qed
    \end{proof}
    \begin{claim} \label{claim:square_permutation_to_square_binary}
        If $\pi$ is a square permutation avoiding $213$ and $231$,
        $\PERMTOBIN(\pi)$ is a square binary word.
    \end{claim}
    \begin{proof}[of Claim~\ref{claim:square_permutation_to_square_binary}]
        Let $\pi$ be a square permutation avoiding $213$ and $231$. By
        Claim~\ref{claim:binary_to_permutation_avoiding}, $\pi$ is in
        the image of $\BINTOPERM$ and hence, $u = \PERMTOBIN(\pi)$ is a
        well-defined binary word. Since $\pi$ is a square permutation,
        there are two disjoint sets $P_1$ and $P_2$ of indexes of letters
        of $\pi$ such that $\pi_{|P_1}$ and $\pi_{|P_2}$ are
        order-isomorphic. This implies, by the definitions of $\BINTOPERM$
        and $\PERMTOBIN$, that $u_{|P_1} = u_{|P_2}$, showing that $u$
        is a square binary word.
        \qed
    \end{proof}
\end{proof}
\medskip

The number of square binary words is Sequence A191755 of~\cite{Slo}
beginning by
\begin{equation}
1, 0, 2, 0, 6, 0, 22, 0, 82, 0, 320, 0, 1268, 0, 5102, 0, 020632.
\end{equation}
By Proposition~\ref{prop:bijection_binary_to_permutations_squares}, this
is also the sequence enumerating square permutations avoiding $213$
and $231$.
\medskip

@@

%%
%% ---- Algebraic issues ----
%%
\section{Algebraic issues}
\label{section:Algebraic issues}
The aim of this section is to establish some of properties of the
shuffle product of permutations $\SHUFFLE$. It is worth to note that, as
we will see, algebraic properties of the unshuffling coproduct $\Delta$
of permutations defined in
Section~\ref{section:Shuffle product on permutations} lead to
combinatorial properties of $\SHUFFLE$.
\medskip

@@

    Finally, to prove the commutativity of $\SHUFFLE$, we shall show
    that $\Delta$ is cocommutative, that is for any permutation $\pi$,
    if in the expansion of $\Delta(\pi)$ there is a tensor
    $\sigma \otimes \nu$ with a coefficient $\lambda$, there is in the
    same expansion the tensor $\nu \otimes \sigma$ with the same
    coefficient $\lambda$. Clearly, a product is commutative if and only
    if its dual coproduct is cocommutative. Now, from the
    definition~\eqref{equ:unshuffling_coproduct} of $\Delta$, one
    observes that if the pair $(P_1, P_2)$ of subsets of $[|\pi|]$
    contributes to the coefficient of
    $\STD\left(\pi_{|P_1}\right) \otimes \STD\left(\pi_{|P_2}\right)$,
    the pair $(P_2, P_1)$ contributes to the coefficient of
    $\STD\left(\pi_{|P_2}\right) \otimes \STD\left(\pi_{|P_1}\right)$.
    This shows that $\Delta$ is cocommutative and hence, that $\SHUFFLE$
    is commutative.
    \qed
\end{proof}
\medskip

Proposition~\ref{prop:shuffle_associative_commutative} implies that
$\QQ[S]$ endowed with the unshuffling coproduct $\Delta$ is a
coassociative cocommutative coalgebra, or in an equivalent way, that
$\QQ[S]$ endowed with the shuffle product $\SHUFFLE$ is an associative
commutative algebra.
\medskip

\begin{lemma} \label{lem:endomorphisms}
    The three linear maps
    \begin{equation}
        \phi_1, \phi_2, \phi_3 : \QQ[S] \to \QQ[S]
    \end{equation}
    linearly sending a permutation $\pi$ to, respectively,
    $\widetilde{\pi}$, $\bar \pi$, and $\pi^{-1}$ are endomorphims of
    associative algebras.
\end{lemma}
\begin{proof}[of Lemma~\ref{lem:endomorphisms}]
    To prove, for $j = 1, 2, 3$, that $\phi_j$ is a morphism of
    associative algebras, we have to prove that for all permutations
    $\sigma$ and $\nu$,
    \begin{equation}
        \phi_j(\sigma \SHUFFLE \nu) =
        \phi_j(\sigma) \SHUFFLE \phi_j(\nu).
    \end{equation}
    We shall proceed here as we have done during the proof of
    Proposition~\ref{prop:shuffle_associative_commutative} by showing
    that $\phi_j$ is a morphism of coalgebras, that is
    \begin{equation}
        \Delta \phi_j = (\phi_j \otimes \phi_j) \Delta.
    \end{equation}
    In the sequel, $\pi$ is a permutation.
    \smallskip

    Finally, for any permutation $\tau$, if $P$ is a set of indexes of
    letters of $\tau$, we denote by $P_\tau^{-1}$ the set
    $\{\tau(i) : i \in P\}$. Since the map sending a subset $P$ of
    $[|\pi|]$ to $P_\pi^{-1}$ is a bijection, and since
    $\STD\left(\pi_{|P}\right)^{-1} = \STD\left(\pi^{-1}_{|P_\pi^{-1}}\right)$,
    we have
    \begin{equation} \begin{split}
        \Delta(\phi_3(\pi))
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \STD\left(\phi_3(\pi)_{|P_1}\right)
        \otimes \STD\left(\phi_3(\pi)_{|P_2}\right) \\
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \STD\left(\pi^{-1}_{|P_1}\right)
        \otimes \STD\left(\pi^{-1}_{|P_2}\right) \\
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \STD\left(\pi^{-1}_{|{P_1}_\pi^{-1}}\right)
        \otimes \STD\left(\pi^{-1}_{|{P_2}_\pi^{-1}}\right) \\
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \STD\left(\pi_{|P_1}\right)^{-1}
        \otimes \STD\left(\pi_{|P_2}\right)^{-1} \\
        & = \sum_{P_1 \sqcup P_2 = [|\pi|]}
        \phi_3\left(\STD\left(\pi_{|P_1}\right)\right)
        \otimes \phi_3\left(\STD\left(\pi_{|P_2}\right)\right) \\
        & = (\phi_3 \otimes \phi_3) \Delta(\pi).
    \end{split} \end{equation}
    This shows that $\phi_3$ is a morphism of coalgebras and hence, that
    $\phi_3$ is a morphism of associative algebras.
    \qed
\end{proof}
\medskip

We now use the algebraic properties of $\SHUFFLE$ exhibited by
Lemma~\ref{lem:endomorphisms} to obtain combinatorial properties
of square permutations.
\medskip

\begin{proposition} \label{prop:square_stability}
    Let $\pi$ be a square permutation and $\sigma$ be a square root of
    $\pi$. Then,
    \begin{enumerate}[label={\it (\roman*)},fullwidth]
        \item \label{item:square_stability_1}
        the permutation $\widetilde{\pi}$ is a square and
        $\widetilde{\sigma}$ is one of its square roots;
        \item \label{item:square_stability_2}
        the permutation $\bar \pi$ is a square and $\bar \sigma$ is one of
        its square roots;
        \item \label{item:square_stability_3}
        the permutation $\pi^{-1}$ is a square and $\sigma^{-1}$ is one of
        its square roots.
    \end{enumerate}
\end{proposition}
\begin{proof}[of Proposition~\ref{prop:square_stability}]
    All statements~\ref{item:square_stability_1},
    \ref{item:square_stability_2}, and~\ref{item:square_stability_3} are
    consequences of Lemma~\ref{lem:endomorphisms}. Indeed,
    since $\pi$ is a square permutation and $\sigma$ is a square root of
    $\pi$, by definition, $\pi$ appears in the product
    $\sigma \SHUFFLE \sigma$. Now, by Lemma~\ref{lem:endomorphisms},
    for any $j = 1, 2, 3$, since $\phi_j$ is a morphism of associative
    algebras from $\QQ[S]$ to $\QQ[S]$, $\phi_j$ commutes with the
    shuffle product of permutations $\SHUFFLE$. Hence, in particular,
    one has
    \begin{equation}
        \phi_j(\sigma \SHUFFLE \sigma) =
        \phi_j(\sigma) \SHUFFLE \phi_j(\sigma).
    \end{equation}
    Then, since $\pi$ appears in $\sigma \SHUFFLE \sigma$, $\phi_j(\pi)$
    appears in $\phi_j(\sigma \SHUFFLE \sigma)$ and appears also in
    $\phi_j(\sigma) \SHUFFLE \phi_j(\sigma)$. This shows that
    $\phi_j(\sigma)$ is a square root of $\phi_j(\pi)$ and
    implies~\ref{item:square_stability_1}, \ref{item:square_stability_2},
    and~\ref{item:square_stability_3}.
    \qed
\end{proof}
\medskip

>>>>

\begin{abstract}
    A permutation is said to be a square if it can be obtained by
    shuffling two order-isomorphic patterns. The definition is intended
    to be the natural counterpart to the ordinary shuffle of words and
    languages. In this paper, we tackle the problem of recognizing square
    permutations from both the point of view of algebra and algorithmic.
    On the one hand, we present some algebraic and combinatorial
    properties of the shuffle product of permutations. We follow an
    unusual line consisting in defining the shuffle of permutations by
    means of an unshuffling operator, known as a coproduct. This
    strategy allows to obtain easy proofs for algebraic and combinatorial
    properties of our shuffle product. We besides exhibit a bijection
    between square $(213,231)$-avoiding permutations and square binary
    words.
    On the other hand,
    using a containment-free perfect matching criterion,
    we prove that recognizing square permutations is \NPC. 
\end{abstract}

@@

The purpose of this section is to define a shuffle product $\SHUFFLE$
on permutations. To define this shuffle product, we shall first define
a coproduct denoted by $\Delta$, enabling to unshuffle permutations.
By duality, $\Delta$ implies the definition of $\SHUFFLE$. The reason
why we need to pass by the definition of $\Delta$ to define $\SHUFFLE$
is justified by the fact that a lot of properties of $\SHUFFLE$ depend
of properties of $\Delta$ and that this strategy allows to write concise
and clear proofs of them. We invite the reader unfamiliar with the
concepts of coproduct and duality to consult~\cite{Joni:Rota:1979}
or~\cite{Grinberg:Reiner:2014}.

Let use denote by $\QQ[S]$ the linear span of all permutations. We
define a linear coproduct $\Delta$ on $\QQ[S]$ in the following way. For
any permutation $\pi$, we set
\begin{equation} \label{equ:unshuffling_coproduct}
    \Delta(\pi) =
    \sum_{P_1 \sqcup P_2 = [|\pi|]}
    \STD\left(\pi_{|P_1}\right) \otimes \STD\left(\pi_{|P_2}\right).
\end{equation}
We call $\Delta$ the {\em unshuffling coproduct of permutations}. For
instance,
\begin{equation}
    \Delta(213) =
    \epsilon \otimes 213 + 2. 1 \otimes 12 +
    1 \otimes 21 + 2. 12 \otimes 1 + 21 \otimes 1 + 213 \otimes \epsilon,
\end{equation}
\begin{equation}
    \Delta(1234) =
    \epsilon \otimes 1234 + 4. 1 \otimes 123 + 6. 12 \otimes 12 +
    4. 123 \otimes 1 + 1234 \otimes \epsilon,
\end{equation}
\begin{equation}\begin{split} \label{equ:example_unshuffling_coproduct}
    \Delta(1432) & =
    \epsilon \otimes 1432 + 3. {\bf 1 \otimes 132} + 1 \otimes 321 +
    3. 12 \otimes 21 \\ & + 3. 21 \otimes 12 + 3. 132 \otimes 1 +
    321 \otimes 1 + 1432 \otimes \epsilon.
\end{split}\end{equation}
Observe that the coefficient of the tensor $1 \otimes 132$ is $3$
in~\eqref{equ:example_unshuffling_coproduct} because there are exactly
three ways to extract from the permutation $1432$ two disjoint subwords
respectively order-isomorphic to the permutations $1$ and $132$.

As announced, let us now use $\Delta$ to define a shuffle product on
permutations. As any coproduct, $\Delta$ leads to the definition of a
product obtained by duality in the following way.
From~\eqref{equ:unshuffling_coproduct}, for any permutation $\pi$, we
have
\begin{equation} \label{equ:schematic_coproduct}
    \Delta(\pi) =
    \sum_{\sigma, \nu \in S} \lambda_{\sigma, \nu}^\pi \;
    \sigma \otimes \nu,
\end{equation}
where the $\lambda_{\sigma, \nu}^\pi$ are nonnegative integers. Now,
by definition of duality, the dual product of $\Delta$, denoted by
$\SHUFFLE$, is a linear binary product on $\QQ[S]$. It satisfies, for
any permutations $\sigma$ and $\nu$,
\begin{equation}
    \sigma \SHUFFLE \nu =
    \sum_{\pi \in S}
    \lambda_{\sigma, \nu}^\pi \; \pi,
\end{equation}
where the coefficients $\lambda_{\sigma, \nu}^\pi$ are the ones
of~\eqref{equ:schematic_coproduct}. We call $\SHUFFLE$ the
{\em shuffle product of permutations}. For instance,
\begin{equation}\begin{split} \label{equ:example_shuffle_product}
    12 \SHUFFLE 21 & =
    1243 + 1324 + 2.1342 + 2.1423 + 3. {\bf 1432} + 2134 + 2.2314 \\
    & + 3.2341 + 2413 + 2.2431 + 2.3124 + 3142 + 3.3214 + 2.3241 \\
    & + 3421 + 3.4123 + 2.4132 + 2.4213 + 4231 + 4312.
\end{split}\end{equation}
Observe that the coefficient $3$ of the permutation $1432$
in~\eqref{equ:example_shuffle_product} comes from the fact that the
coefficient of the tensor $12 \otimes 21$ is $3$
in~\eqref{equ:example_unshuffling_coproduct}.

Intuitively, this product shuffles the values and the positions of the
letters of the permutations. As first and easy properties of $\SHUFFLE$,
one can observe that the empty permutation $\epsilon$ is a unit for
$\SHUFFLE$ and that this product is graded by the sizes of the
permutations ({\em i.e.}, the product of a permutation of size $n$ with
a permutation of size $m$ produces a sum of permutations of size $n + m$).
Notice that there are other shuffle products on
permutations~\cite{DHT:IJAC:2002}: the {\em shifted shuffle} and the
{\em convolution product}. Our shuffle product is different from these
last two.

We say that a permutation $\pi$ {\em appears} in the shuffle
$\sigma \SHUFFLE \nu$ of two permutations $\sigma$ and $\nu$ if the
coefficient $\lambda_{\sigma, \nu}^\pi$ defined above is different from
zero. In a more combinatorial way, this is equivalent to say that there
are two sets $P_1$ and $P_2$ of disjoints indexes of letters of $\pi$
satisfying $P_1 \sqcup P_2 = [|\pi|]$ such that the subword $\pi_{|P_1}$
is order-isomorphic to $\sigma$ and the subword $\pi_{|P_2}$ is
order-isomorphic to $\nu$.

@@

\begin{tabular}{c|c|c}
    Size $0$ \, & Size $2$ & Size $4$ \\ \hline
    \multirow{2}{*}{\, $\epsilon$} &
    \multirow{2}{*}{\, $12$, $21$ \,} &
    $1234$, $1243$, $1423$, $1324$, $1342$, $4132$, $3124$, $3142$,
    $3412$, $4312$, \\
    & & $2134$, $2143$, $2413$, $4213$, $2314$, $2431$, $4231$, $3241$,
    $3421$, $4321$
\end{tabular}

@@

%%
%% ---- Binary square words and permutations ----
%%
\section{Binary square words and permutations}
\label{section:Binary square words and permutations}
In this section, we shall show that the square binary words are in
one-to-one correspondence with square permutations avoiding some
patterns. This property establishes a link between the shuffle of binary
words and our shuffle of permutations and allows to obtain a new
description of square binary words.

Let $u$ be a binary word of length $n$ with $k$ occurrences of $0$.
We denote by $\BINTOPERM$ the map sending any such word $u$ to the
permutation obtained by replacing from left to right each occurrence of
$0$ in $u$ by $1$, $2$, \dots, $k$, and from right to left each
occurrence of $1$ in $u$ by $k + 1$, $k + 2$, \dots, $n$. For instance,
\begin{equation}
    \BINTOPERM({\bf 1}00{\bf 1}0{\bf 1}{\bf 1}0{\bf 1}000) =
    {\bf C} 12 {\bf B} 3 {\bf A} {\bf 9} 4 {\bf 8} 567.
\end{equation}
Observe that for any nonempty permutation $\pi$ in the image of
$\BINTOPERM$, there is exactly one binary word $u$ such that
$\BINTOPERM(u0) = \BINTOPERM(u1) = \pi$. In support of this observation,
when $\pi$ has an even size, we denote by $\PERMTOBIN(\pi)$ the word $ua$
such that $|ua|_0$ and $|ua|_1$ are both even, where $a \in \{0, 1\}$.

\begin{proposition} \label{prop:bijection_binary_to_permutations_squares}
    For any $n \geq 0$, the map $\BINTOPERM$ restricted to the set of
    square binary words of length $2n$ is a bijection between this last
    set and the set of square permutations of size $2n$ avoiding the
    patterns $213$ and $231$.
\end{proposition}
\begin{proof}
    [of Proposition~\ref{prop:bijection_binary_to_permutations_squares}]
    The statement of the proposition is a consequence of the
    following claims implying that $\PERMTOBIN$ is the inverse
    map of $\BINTOPERM$ over the set of square binary words.
    \begin{claim} \label{claim:binary_to_permutation_avoiding}
        The image of $\BINTOPERM$ is the set of all permutations
        avoiding $213$ and $231$.
    \end{claim}

    \begin{claim} \label{claim:square_binary_to_square_permutation}
        If $u$ is a square binary word, $\BINTOPERM(u)$ is a square
        permutation.
    \end{claim}

    \begin{claim} \label{claim:square_permutation_to_square_binary}
        If $\pi$ is a square permutation avoiding $213$ and $231$,
        $\PERMTOBIN(\pi)$ is a square binary word.
    \end{claim}
\end{proof}

The number of square binary words is Sequence A191755 of~\cite{Slo}
beginning by
\begin{equation}
1, 0, 2, 0, 6, 0, 22, 0, 82, 0, 320, 0, 1268, 0, 5102, 0, 020632.
\end{equation}
By Proposition~\ref{prop:bijection_binary_to_permutations_squares}, this
is also the sequence enumerating square permutations avoiding $213$
and $231$.

@@

%%
%% ---- Algebraic issues ----
%%
\section{Algebraic issues}
\label{section:Algebraic issues}
The aim of this section is to establish some of properties of the
shuffle product of permutations $\SHUFFLE$. It is worth to note that, as
we will see, algebraic properties of the unshuffling coproduct $\Delta$
of permutations defined in
Section~\ref{section:Shuffle product on permutations} lead to
combinatorial properties of $\SHUFFLE$.

@@

    Finally, to prove the commutativity of $\SHUFFLE$, we shall show
    that $\Delta$ is cocommutative, that is for any permutation $\pi$,
    if in the expansion of $\Delta(\pi)$ there is a tensor
    $\sigma \otimes \nu$ with a coefficient $\lambda$, there is in the
    same expansion the tensor $\nu \otimes \sigma$ with the same
    coefficient $\lambda$. Clearly, a product is commutative if and only
    if its dual coproduct is cocommutative. Now, from the
    definition~\eqref{equ:unshuffling_coproduct} of $\Delta$, one
    observes that if the pair $(P_1, P_2)$ of subsets of $[|\pi|]$
    contributes to the coefficient of
    $\STD\left(\pi_{|P_1}\right) \otimes \STD\left(\pi_{|P_2}\right)$,
    the pair $(P_2, P_1)$ contributes to the coefficient of
    $\STD\left(\pi_{|P_2}\right) \otimes \STD\left(\pi_{|P_1}\right)$.
    This shows that $\Delta$ is cocommutative and hence, that $\SHUFFLE$
    is commutative.
    \qed
\end{proof}

Proposition~\ref{prop:shuffle_associative_commutative} implies that
$\QQ[S]$ endowed with the unshuffling coproduct $\Delta$ is a
coassociative cocommutative coalgebra, or in an equivalent way, that
$\QQ[S]$ endowed with the shuffle product $\SHUFFLE$ is an associative
commutative algebra.

\begin{lemma} \label{lem:endomorphisms}
    The three linear maps
    \begin{equation}
        \phi_1, \phi_2, \phi_3 : \QQ[S] \to \QQ[S]
    \end{equation}
    linearly sending a permutation $\pi$ to, respectively,
    $\widetilde{\pi}$, $\bar \pi$, and $\pi^{-1}$ are endomorphims of
    associative algebras.
\end{lemma}

We now use the algebraic properties of $\SHUFFLE$ exhibited by
Lemma~\ref{lem:endomorphisms} to obtain combinatorial properties
of square permutations.

\begin{proposition} \label{prop:square_stability}
    Let $\pi$ be a square permutation and $\sigma$ be a square root of
    $\pi$. Then,
    \begin{enumerate}[label={\it (\roman*)},fullwidth]
        \item \label{item:square_stability_1}
        the permutation $\widetilde{\pi}$ is a square and
        $\widetilde{\sigma}$ is one of its square roots;
        \item \label{item:square_stability_2}
        the permutation $\bar \pi$ is a square and $\bar \sigma$ is one of
        its square roots;
        \item \label{item:square_stability_3}
        the permutation $\pi^{-1}$ is a square and $\sigma^{-1}$ is one of
        its square roots.
    \end{enumerate}
\end{proposition}
\begin{proof}[of Proposition~\ref{prop:square_stability}]
    All statements~\ref{item:square_stability_1},
    \ref{item:square_stability_2}, and~\ref{item:square_stability_3} are
    consequences of Lemma~\ref{lem:endomorphisms}. Indeed,
    since $\pi$ is a square permutation and $\sigma$ is a square root of
    $\pi$, by definition, $\pi$ appears in the product
    $\sigma \SHUFFLE \sigma$. Now, by Lemma~\ref{lem:endomorphisms},
    for any $j = 1, 2, 3$, since $\phi_j$ is a morphism of associative
    algebras from $\QQ[S]$ to $\QQ[S]$, $\phi_j$ commutes with the
    shuffle product of permutations $\SHUFFLE$. Hence, in particular,
    one has
    \begin{equation}
        \phi_j(\sigma \SHUFFLE \sigma) =
        \phi_j(\sigma) \SHUFFLE \phi_j(\sigma).
    \end{equation}
    Then, since $\pi$ appears in $\sigma \SHUFFLE \sigma$, $\phi_j(\pi)$
    appears in $\phi_j(\sigma \SHUFFLE \sigma)$ and appears also in
    $\phi_j(\sigma) \SHUFFLE \phi_j(\sigma)$. This shows that
    $\phi_j(\sigma)$ is a square root of $\phi_j(\pi)$ and
    implies~\ref{item:square_stability_1}, \ref{item:square_stability_2},
    and~\ref{item:square_stability_3}.
    \qed
\end{proof}

Let us make an observation about Wilf-equivalence classes of
permutations restrained on square permutations. The
well-known~\cite{Simion:Schmidt:EJC:1985} fact that there is a single
Wilf-equivalence class of permutations of size $3$ together with
Proposition~\ref{prop:square_stability} imply that $123$ and $321$ are
in the same Wilf-equivalence class of square permutations, and that
$132$, $213$, $231$, and $312$ are in the same Wilf-equivalence class of
square permutations. Computer experiments show us that there are two
Wilf-equivalence classes of square permutations of size $3$. Indeed, the
number of square permutations avoiding $123$ begins by
\begin{equation} \label{equ:sequence_square_123}
    1, 0, 2, 0, 12, 0, 118, 0, 1218, 0, 14272,
\end{equation}
while the number of square permutations avoiding $132$ begins by
\begin{equation} \label{equ:sequence_square_132}
    1, 0, 2, 0, 11, 0, 84, 0, 743, 0, 7108.
\end{equation}

Besides, an other consequence of Proposition~\ref{prop:square_stability}
is that its makes sense to enumerate the sets of square permutations
quotiented by the operations of mirror image, complementary, and
inverse. The sequence enumerating these sets begins by
\begin{equation} \label{equ:sequence_square_classes}
    1, 0, 1, 0, 6, 0, 81, 0, 2774, 0, 162945.
\end{equation}

All Sequences~\eqref{equ:sequence_square_123}, \eqref{equ:sequence_square_132},
and~\eqref{equ:sequence_square_classes} (and their subsequences obtained
by removing the $0$s) are for the time being not listed in~\cite{Slo}.
